Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-09 paper/paper.tex
edit on reference, plot layout & captions, acknowleedgements

Visualisation is widely used in exploratory data analysis (REF). Presenting information in a graphical format often allows people to uncover information they would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper combines data visualization with optimization methods. It aims at bring visualization tools into optimization problems to assist users to better understand the performance of their optimization methods in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (\citet{kelley1999iterative}). The unconstrained problem can be formulated as a minimization (or maximization) problem such as
\(min_{x} f(x)\) where \(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with certain properties defined in an \(L^p\) space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions are introduced via a set of functions that can be convex or non-convex: \(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and hence the problem can be written as
\(min_{x} f(x)\) \emph{subject to} \(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (\citet{buja2005computational}) which is an exploratory data analysis tool that is defined to detect \emph{interesting structures} or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (cook2008grand). The target of the optimisation is to identify the most \emph{interesting} low-dimensional views of the data given by a corresponding projection matrix. The most \emph{interesting} structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (\citet{cook1993projection}).

The optimization challenges encountered in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple maxima (local and global), the trade off between computational burden and proximity to the maxima, dealing with noisy objective functions that might be non smooth and non diferentiable (\citet{jones1998efficient}). Those are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

In gradient descent methods the goal is to find the minimum of an objective function while in gradient ascent methods the interest is in the maximum. The minimum is searched for by taken steps proportional to the negative gradient or of the function at the evaluation point (analogously positive gradient for gradient ascent methods), that is \(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or the learning rate (\citet{fletcher2013practical}). The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned (\citet{trefethen1997numerical}). In order to solve some of those problems, line minimization or line search methods were proposed (\citet{shi2004convergence}) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for each iteration where the objective is to find the value of \(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) (this step is called the line search). Then move on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction \(S_{k}\), and on approximations of the step size \(\alpha_k\) together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the existence of the objective function derivatives and on the complexity of the constraints. Because of that, the second big class of optimization problems look at derivative free methods where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology (REF), stochastic approximation (REF),
random search (REF) and heuristic methods like tabu search and genetic algorithms (REF). Later, we will present an optimization algorithm within the random search class, namely simulated annealing for optimization with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions both in gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both general purpose as well as tasks specifics solvers. The most prominent one within the general solvers is \emph{optim} in (REF-stats package) which have functions for gradient based and derivative free methods. Other packages having similar functionalities include \emph{nloptr} (REF), XX.

@@

Mathematical details of the geodesic interpolation can be found in \citet{buja2005computational}. Figure \ref{fig:tour-path} shows a sketch of the tour path. The blue frames are produced by the projection pursuit optimisation algorithm, and the white frames interpolate between them. The tour method has been implemented in the \texttt{tourr} package in R, available on the Comprehensive R Archive Network at \url{https://cran.r-project.org/web/packages/tourr/} \citep{wickham2011tourrpackage}.

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth,height=0.3\textheight]{/Users/pmen0008/Desktop/paper-tour-vis/img/tour_path}

@@

\caption{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}

@@

The optimisation problem in the tour context is stated as follows:
\ldots.
.Given a randomly generated starting basis \(\mathbf{A}_1\), projection pursuit finds the final projection basis \(\mathbf{A}_T\) that satisfies the following optimisation problem:

@@

where \(I_d\) is the \(d\)-dimensional identity matrix and the constraint requires the projection bases \(\mathbf{A}\) to be orthogonal matrices.

@@

where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is no output in iteration \(J + 1\) since the fact that it is the last iteration means that the optimiser cannot find a better basis and the algorithm terminates. In this notation, the final basis found is \(A_T\) with the highest index value \(I_T\).

The data structure constructed above meets the tidy data principle \citep{wickham2014tidy} that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by well-developed packages such as dplyr\citep{dplyr} and ggplot2\citep{ggplot2}.

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-search-1}

@@

\caption{A comparison of \texttt{search\_better} with different values of the parameter \texttt{max\_tries}. A six-variable dataset \texttt{boa6} is used with \texttt{holes} index on a 2D problem. \texttt{max\_tries\ =\ 500} allows the optimiser to find better basis with higher index value at iteration six.}\label{fig:toy-search}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-interp-1}

@@

\caption{The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.}\label{fig:toy-interp}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca-1}

@@

\caption{1D projection on the 5-variable dataset \texttt{boa5} with two optimisers: \texttt{search\_better} and \texttt{search\_geodesic}. The yellow point corresponds to the theoretical best basis {[}0, 1, 0, 0, 0{]} with \texttt{V2} being the only non-normal variable in the dataset. The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \texttt{search\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.}\label{fig:toy-pca}
\end{figure}

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases \(\mathbf{A}_{p \times d}\) live on the surface of a \(p \times d\) dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space (\(d > 3\)), an easier alternative is to view the reduced space via some dimension reduction methods, i.e.~principal component analysis. To better perceive the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package \texttt{geozoo} and PCA is conducted on both the bases and the points on the surface of the sphere.

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca-animated-1}

@@

\caption{A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.}\label{fig:toy-pca-animated}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-tour-1}

@@

\caption{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.}\label{fig:toy-tour}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/interruption-1}

@@

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_better}. A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.}\label{fig:interruption}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/polish-1}

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish.}\label{fig:polish}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/noisy-better-geo-1}

\caption{One-D projection on \texttt{boa5} data with noisy index \texttt{norm\_kol} optimised by \texttt{search\_geodesic} and \texttt{search\_better}. The grey dashed line represents the index value of the theoretical best basis. \texttt{search\_geodesic} fails to optimise the noisy index while \texttt{search\_better} has made reasonable improvements to reach the index value close to the one of the theoretical best basis.}\label{fig:noisy-better-geo}
\end{figure}

@@

Figure \ref{fig:noisy-better-geo} compares the tracing plot of two optimisers: \texttt{search\_geodesic} and \texttt{search\_better}. This time, the interpolated path is no longer smooth when using either algorithm and \texttt{search\_geodesic} fails to optimise this index with barely any improvement of the index value. On the other hand, \texttt{search\_better} is doing relatively well. With the theoretical best basis {[}0, 1, 0, 0, 0{]} producing an index value of 0.182, \texttt{search\_better} finds the final basis {[} 0.0376, -0.9916, -0.0581, -0.0831, 0.0716{]} with an index value of 0.165. A further polish step will give a marginal improvement of index value to 0.175 with a basis of {[} 0.0223, -0.9965, -0.0352, -0.0591, 0.0418{]}. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-better-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better} with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when \texttt{V2} is found and the global maximum where \texttt{V7} is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.}\label{fig:kol-better}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-random-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with the same seeds. \texttt{search\_better\_random} has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum.}\label{fig:kol-random}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-random-tuned-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7.}\label{fig:kol-random-tuned}
\end{figure}

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is \texttt{boa6} where \texttt{x2} and \texttt{x7} are informative. The two theoretical best bases are {[}0, 1, 0, 0, 0, 0{]} and {[}0, 0, 1, 0, 0, 0{]} with index value 0.182 and 0.235, respectively. Hence, the global maximum happens when variable \texttt{x7} is found. Simulation is done with 20 randomly generated seeds on two optimisers: \texttt{search\_better} (Figure \ref{fig:kol-better}) and \texttt{search\_better\_random} (Figure \ref{fig:kol-random} and Figure \ref{fig:kol-random-tuned}). Compared to Figure \ref{fig:kol-better} and \ref{fig:kol-random}, Figure \ref{fig:kol-random-tuned} further increases the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7 to enlarge the search space. The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

@@

Most of the work in this project has been translated into code in two packages: the collection of the data object is implemented in the existing \texttt{tourr} package; manipulation and visualisation of the data object are implemented in the new \texttt{ferrn} package. Equipped with handy tools to diagnose the performance of optimisers, future work can extend the diagnostics to a wider range of index functions, i.e.~scagnostics, association, and information index \citep{laa2020using} and understand how the optimisers behave for index functions with different structures.

>>>>

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory}. Presenting information in a graphical
format often allows people to uncover information they would otherwise
not be aware of and provides a more comprehensive understanding of the
problem at hand. The work presented in this paper combines data
visualization with optimization methods. It aims at bring visualization
tools into optimization problems to assist users to better understand
the performance of their optimization methods in practice. This
motivates our work of creating plots to diagnose optimisation algorithms
which in turn can be also used not only to understand optimization
performance but also to understand and compare the behaviour of
different algorithms.

The goal of continuous optimisation is to find the best solution within
the space of all feasible solutions where typically the best solution is
decided by an objective function. Broadly speaking optimization can be
unconstrained or constrained \citep{kelley1999iterative}. The
unconstrained problem can be formulated as a minimization (or
maximization) problem such as \(\min_{x} f(x)\) where
\(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with
certain properties defined in an \(L^p\) space. In this case, solutions
rely on gradient descent or ascent methods. In the constrained
optimization problem additional restrictions are introduced via a set of
functions that can be convex or non-convex:
\(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and
hence the problem can be written as \(\min_{x} f(x)\) \emph{subject to}
\(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex
optimization methods including linear and quadratic programming can be
used.

The motivation of this paper is the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational} which is an
exploratory data analysis tool that is defined to detect
\emph{interesting structures} or features in high-dimensional data
through a set of lower-dimensional projections that cover the entire
high dimensional space using interpolation methods called tours
\citep{cook2008grand}. The target of the optimisation is to identify the
most \emph{interesting} low-dimensional views of the data given by a
corresponding projection matrix. The most \emph{interesting} structures
are formally defined by a function of projections, called index function
which is optimized to uncover the most reveling structures in a high
dimensional space \citep{cook1993projection}.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples of
those include the existence of multiple maxima (local and global), the
trade off between computational burden and proximity to the maxima,
dealing with noisy objective functions that might be non smooth and non
diferentiable \citep{jones1998efficient}. Those are not unique to this
context and therefore the visualization tools and optimization methods
presented in this paper can be easily apply to any other optimization
problems.

@@

In gradient descent methods the goal is to find the minimum of an
objective function while in gradient ascent methods the interest is in
the maximum. The minimum is searched for by taken steps proportional to
the negative gradient or of the function at the evaluation point
(analogously positive gradient for gradient ascent methods), that is
\(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or
the learning rate \citep{fletcher2013practical}. The step size is the
key in the success of the optimization problem as inadequate step sizes
might lead to a failure in convergence. These methods are known to be
relative slow and can have convergence issues when problems are poorly
conditioned \citep{trefethen1997numerical}. In order to solve some of
those problems, line minimization or line search methods were proposed
\citep{shi2004convergence} where the step size is chosen in such a way
that convergence is ensured. In line search methods users are required
to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for
each iteration where the objective is to find the value of
\(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with
respect to \(\alpha_k\) (this step is called the line search). Then move
on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and
repeat the process until the desire convergence is reached. Modern
development of line search methods focuses on improving the computation
of the searching direction \(S_{k}\), and on approximations of the step
size \(\alpha_k\) together with reducing the computational storing
burden of the intermediate computations while catering for practical
optimisation problems. In the next section two line search methods for
the optimisation used in the projection pursuit guided tour will be
presented.

Although gradient optimization methods are very popular, they rely on
the existence of the objective function derivatives and on the
complexity of the constraints. Because of that, the second big class of
optimization problems look at derivative free methods where the emphasis
is on finding, in most cases, a near optimal solution. Examples of those
include response surface methodology \citep{box1951experimental},
stochastic approximation \citep{robbins1951stochastic}, random search
\citep{fu2015handbook} and heuristic methods
\citep{sorensen2013metaheuristics}. Later, we will present an
optimization algorithm within the random search class, namely simulated
annealing for optimization with the guided tour. Moreover, we will
introduce an algorithm to further improve optimization solutions both in
gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both
general purpose as well as tasks specifics solvers. The most prominent
one within the general solvers is \texttt{optim()} in the package
\CRANpkg{stats}\citep{stats} which have functions for gradient based and
derivative free methods. Other packages having similar functionalities
include \CRANpkg{nloptr}.

@@

Mathematical details of the geodesic interpolation can be found in
\citet{buja2005computational}. Figure \ref{fig:tour-path} shows a sketch
of the tour path. The blue frames are produced by the projection pursuit
optimisation algorithm, and the white frames interpolate between them.
The tour method has been implemented in the R package \CRANpkg{tourr}
\citep{tourr}, available on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=0.5\linewidth,height=0.25\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour_path}

@@

\caption[Each square (frame) represents the projected data with a corresponding basis]{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}
\end{Schunk}

@@

pa The optimisation problem in the tour context is stated as follows:
\ldots. .Given a randomly generated starting basis \(\mathbf{A}_1\),
projection pursuit finds the final projection basis \(\mathbf{A}_T\)
that satisfies the following optimisation problem:

@@

\noindent where \(I_d\) is the \(d\)-dimensional identity matrix and the
constraint requires the projection bases \(\mathbf{A}\) to be orthogonal
matrices.

@@

\noindent where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is
no output in iteration \(J + 1\) since the fact that it is the last
iteration means that the optimiser cannot find a better basis and the
algorithm terminates. In this notation, the final basis found is \(A_T\)
with the highest index value \(I_T\).

The data structure constructed above meets the tidy data principle
\citep{wickham2014tidy} that requires each observation forms a row and
each variable forms a column. With tidy data structure, data wrangling
and visualisation have been significantly simplified by well-developed
packages such as \CRANpkg{dplyr} \citep{dplyr} and \CRANpkg{ggplot2}
\citep{ggplot2}.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

\caption{A comparison of \code{search\_better} with different values of the parameter \code{max\_tries}. A six-variable dataset \code{boa6} is used with holes index on a 2D problem. \code{max\_tries = 500} allows the optimiser to find better basis with higher index value at iteration six.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

\caption[The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index]{The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

\caption{1D projection on the 5-variable dataset \code{boa5} with two optimisers: \code{search\_better} and \code{search\_geodesic}. The yellow point corresponds to the theoretical best basis [0, 1, 0, 0, 0] with `V2` being the only non-normal variable in the dataset.  The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \code{search\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

Apart from checking the progression of an optimiser, another interesting
aspect is to visualise how the search looks like in its parameter space.
Given the orthonormality constraint, the projection bases
\(\mathbf{A}_{p \times d}\) live on the surface of a \(p \times d\)
dimension sphere, where the dimension can easily go over two or three.
While visualising the search paths on the original high dimensional
sphere would require skills for the viewers to perceive rotation of
geometry in higher dimensional space (\(d > 3\)), an easier alternative
is to view the reduced space via some dimension reduction methods,
i.e.~principal component analysis. To better perceive the search path as
an embedding of a hollow sphere, random points on the high dimensional
sphere is generated using the package \CRANpkg{geozoo} \citep{geozoo}
and PCA is conducted on both the bases and the points on the surface of
the sphere.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

\caption[A selected number of frames from the animated PCA plot]{A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

\caption[A selected number of frames from the tour animation for viewing the 5D space of all the projection bases]{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

\caption[Two-D projection on \code{boa6} data with holes index optimised by \code{search\_geodesic}]{Two-D projection on \code{boa6} data with holes index optimised by \code{search\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. }\label{fig:polish}
\end{figure}
\end{Schunk}

@@

@@

Figure \ref{fig:noisy-better-geo} compares the tracing plot of two
optimisers: \texttt{search\_geodesic} and \texttt{search\_better}. This
time, the interpolated path is no longer smooth when using either
algorithm and \texttt{search\_geodesic} fails to optimise this index
with barely any improvement of the index value. On the other hand,
\texttt{search\_better} is doing relatively well. With the theoretical
best basis {[}0, 1, 0, 0, 0{]} producing an index value of 0.17,
\texttt{search\_better} finds the final basis {[} 0.0376, -0.9916,
-0.0581, -0.0831, 0.0716{]} with an index value of 0.165. A further
polish step will give a marginal improvement of index value to 0.175
with a basis of {[} 0.0223, -0.9965, -0.0352, -0.0591, 0.0418{]}. At
this stage, the difference between the theoretical best and what has
been found is likely due to simulation error since the best possible
basis for a simulated data will be slightly off the theoretical best
basis, which is derived based on the distributional assumptions in
Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-better-1}

@@

\caption[One-D projection of \code{norm\_kol} index on \code{boa6} data optimised by \code{search\_better} with 20 randomly generated seeds]{One-D projection of \code{norm\_kol} index on \code{boa6} data optimised by \code{search\_better} with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when `V2` is found and the global maximum where `V7` is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.}\label{fig:kol-better}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-random-1}

@@

\caption{One-D projection of \code{norm\_kol} index on \code{boa6} data optimised by \code{search\_better\_random} with the same seeds. \code{search\_better\_random} has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum.}\label{fig:kol-random}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-random-tuned-1}

@@

\caption[One-D projection of \code{norm\_kol} index on \code{boa6} data optimised by \code{search\_better\_random} with an larger searching neighbourhood of 0.7]{One-D projection of \code{norm\_kol} index on \code{boa6} data optimised by \code{search\_better\_random} with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter `alpha` from 0.5 to 0.7. }\label{fig:kol-random-tuned}
\end{figure}
\end{Schunk}

The second experiment with the noisy index is to understand how the
optimisers perform when a local maximum is present. The dataset used is
\textbackslash code\{boa6\} where \texttt{x2} and \texttt{x7} are
informative. The two theoretical best bases are {[}0, 1, 0, 0, 0, 0{]}
and {[}0, 0, 1, 0, 0, 0{]} with index value 0.176 and 0.235,
respectively. Hence, the global maximum happens when variable
\texttt{x7} is found. Simulation is done with 20 randomly generated
seeds on two optimisers: \texttt{search\_better} (Figure
\ref{fig:kol-better}) and \texttt{search\_better\_random} (Figure
\ref{fig:kol-random} and Figure \ref{fig:kol-random-tuned}). Compared to
Figure \ref{fig:kol-better} and \ref{fig:kol-random}, Figure
\ref{fig:kol-random-tuned}further increases the neighbourhood parameter
\texttt{alpha} from 0.5 to 0.7 to enlarge the search space. The data
object is collected for each simulation and computation is done to
reduce the bases from its original 6D to a 2D points by principal
component analysis so as to view the search paths.

@@

Most of the work in this project has been translated into code in two
packages: the collection of the data object is implemented in the
existing \CRANpkg{tourr}\citep{tourr} package; manipulation and
visualisation of the data object are implemented in the new \pkg{ferrn}
package. Equipped with handy tools to diagnose the performance of
optimisers, future work can extend the diagnostics to a wider range of
index functions, i.e.~scagnostics, association, and information index
\citep{laa2020using} and understand how the optimisers behave for index
functions with different structures.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This article is created using \CRANpkg{knitr}\citep{knitr} and
\CRANpkg{rmarkdown} \citep{rmarkdown} in R. The source code for
reproducing this paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-tour-vis}.
